---
title: "Friday_Session_Spe28"
author: "Zhidong Chen"
date: "9/28/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set root directory 
require("knitr")
opts_knit$set(root.dir = "~/Documents/AAE636_2018Fall/FridaySessions/2_Sep28")
```

# Today
  1. **Data Frame**
  2. **Writing Functions**
  3. **Plots**

## 1. Data Frame

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# Load packages
library(readxl) 
library(dplyr)
library(psych)
library(pastecs)
library(ggplot2)
library(gridExtra)
```

# ... Read and Write data

Read excel file:

```{r, warning=FALSE}
# read data
house636 <- readxl::read_excel("house636.xlsx", sheet = "Data", col_names = TRUE)
```

Most commonly-used data format: .csv -- comma-separted values files. 
read.csv("path/file-name.csv")
write.csv("path/new-file-name.csv")


### 1.1 Commonly-used functions

R has some built-in data. **mtcars** is one of them. We use **mtcars** for 
today's illustration. 

Type "?mtcars" in your Rstudio console for more information.

```{r}
m <- mtcars 
# work on a copy,  espcially when your data size is big and takes time to load.

str(m) 
# structure of the data

dim(m) # dimentions of the data
ncol(m) # number of columns
nrow(m) # number of rows

head(m)
# quick look at the first couple (6 by default) rows
head(m, n = 15)

tail(m)
tail(m, n = 10)
# last n rows

summary(m)
# summary statisics of the data
```

### 1.2 Substting a Data Frame 

```{r}

# subset by index
m[ ,2]       # 2nd column
m[1, ]       # 1st row
m[2,3]       # 2nd row, 3rd column

# subset by variable name
names(m) # names() list all variable names of your data in order 
m$mpg # data$variable.name

# further subsetting of the mpg column refers to vector subsetting
m$mpg[20] # vector subsetting by index
m$mpg[m$mpg > mean(m$mpg)] # vector subsetting by logial expressions

# How many cars have above-the-average mpg?
length(m$mpg[m$mpg > mean(m$mpg)]) 
# Index of cars with above-the-average mpg
which(m$mpg > mean(m$mpg))
# Names of cars with above-the-average mpg
rownames(m)[m$mpg > mean(m$mpg)]

# subset data frame using combined methods
m[m$mpg > mean(m$mpg), c(1,2,3)]
m[m$mpg > mean(m$mpg), c("mpg","cyl","disp")]


# add a column : manufacturer's suggested retail price (MSRP)
m$MSRP <- 20000 * c(1:nrow(m))
# Often, we focus on log price:
m$lnMSRP <- log(m$MSRP)

# delete a column
m$MSRP <- NULL #or
m <- m[, -12]
```

## 2. Write Functions

Create a summary statistics table with **Minimum, Median, Mean, 
Maximum and Standard Deviation**.

### 2.1 Try to use functions from packages

```{r}

summary(m)
summary(m[, c("mpg", "disp", "hp")])

```

**or**

```{r, message = FALSE}

psych::describe(m)
psych::describe(m[, c("mpg", "disp", "hp")])
pastecs::stat.desc(m)

```

### 2.1 Write your own summary statistics function

Function.Name <- function(Parameters){

BODY OF YOUR FUNCTION

}


```{r}
summary_sd <- function(x) {
        
        c(Min = min(x), Mean = mean(x), 
          Median = median(x), Max = max(x), SD = sd(x))
}

# summary_sd(m) ## error message ...
# type of your input..
summary_sd(m$mpg)

# Too many digits after the decimal point?  
summary_sd <- function(x) {
        
        round(c(Min = min(x), Mean = mean(x), Median = median(x), 
                Max = max(x), SD = sd(x)), 3)
}


# How to apply a function to all of the varialbes in your data?
sapply(m[, c("mpg", "disp", "hp")], summary_sd)
```

Reference:  apply functions -
https://www.r-bloggers.com/apply-lapply-rapply-sapply-functions-in-r/

  

## 3. Plots

### 3.1 Create a scatterplot:

```{r}
plot(x = m$hp, y = m$mpg)

# add a title
plot(x = m$hp, y =m$mpg, main = "Relationship Between Miles/Gallon and Horse Power")

# change labels of the axises
plot(x = m$hp, y =m$mpg, main = "Relationship Between Miles/Gallon and Horse Power",
     xlab = "Horse Power", ylab = "Miles/Gallon")

# change plotting character
plot(x = m$hp, y =m$mpg, main = "Relationship Between Miles/Gallon and Horse Power",
     xlab = "Horse Power", ylab = "Miles/Gallon",
     pch = 16) # ?points

# change color of plot
plot(x = m$hp, y =m$mpg, main = "Relationship Between Miles/Gallon and Horse Power",
     xlab = "Horse Power", ylab = "Miles/Gallon",
     pch = 16,
     col = "purple") # ?points

```
Reference for more formatting: https://www.statmethods.net/advgraphs/parameters.html

### 3.2 Create a histgram:

```{r}
hist(m$mpg, freq = FALSE, main = "Histogram of Miles/Gallon", xlab = "MPG")
hist(m$mpg, freq = TRUE, main = "Histogram of Miles/Gallon", xlab = "MPG")

```

### 3.3 Two graphs in one panel

```{r}
par(mfrow = c(1,2))

plot(x = m$hp, y =m$mpg, main = "Miles/Gallon vs Horse Power",
     xlab = "Horse Power", ylab = "Miles/Gallon",
     pch = 16,
     col = "purple") # ?points

hist(m$mpg, freq = FALSE, main = "Histogram of Miles/Gallon", xlab = "MPG")

par(mfrow = c(1,1))


```

Save you plot
pdf("path/plot-name.pdf")
plot()
dev.off()

### 3.4 Powerful **ggplot2** !!

#### 3.4.1 Scatter plot
```{r}
# install.packages("ggplot2")

# plot mpg vs hp
ggplot(data = m, aes(hp, mpg)) +
        geom_point()


plot.mpg.hp <- ggplot(data = m, aes(hp, mpg)) +
# change color, shape, and size of the points
        geom_point(colour = "purple", shape = 18, size = 3) +
# add title and change axis names
        ggtitle("Miles/Gallon vs Horse Power") +
        xlab("Miles/Gallon") +
        ylab("Horse Power") +
#  add options - move the title to  center, and make it bigger.. nicer!! ...    
        theme(title = element_text(size=12, face = 'bold'),
     plot.title = element_text(hjust = 0.5))

        
```

Good Reference: https://ggplot2.tidyverse.org
List several good places to start your journey of **ggplot2**. 

#### 3.4.2 Histogram 


```{r}
ggplot(data = m, aes(mpg)) +
        geom_histogram(bins = 8, fill = "blue", alpha = .7) +
        # add title and change axis names
        ggtitle("Histogram of Horse Power") +
        xlab("Horse Power") +
        ylab("Count") 

hist.mpg <- ggplot(data = m, aes(mpg)) +
        geom_histogram(bins = 8, fill = "blue", alpha = .7) +
        # add title and change axis names
        ggtitle("Histogram of Horse Power") +
        xlab("Horse Power") +
        ylab("Count") 

ggsave(hist.mpg, filename = "~/Documents/AAE636_2018Fall/FridaySessions/2_Sep28/mpg-hp.pdf",
       device = "pdf")

```

### 3.4.3 Put two graphs in one panel

```{r, message=FALSE}

# install.packages("gridExtra")
gridExtra::grid.arrange(plot.mpg.hp, hist.mpg, ncol=2)
# save your plot to a pdf file
ggsave(filename = "~/Documents/AAE636_2018Fall/FridaySessions/2_Sep28/two-in-one.pdf",
       device = "pdf")

```


