---
title: "Tutorial 3"
author: "Bethany Glinsmann"
date: "3/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set root directory 
require("knitr")
opts_knit$set(root.dir = "~/AAE 770 - math/R_Materials_AAE636_Zhidong/Bethany/")
```

# Today
  1. **Intro to plotting in R**
  2. **Scatterplots**
  3. **Histograms**
  4. **Plot aesthetics**
  5. **Arranging multiple plots**


## 0. Load required packages and data

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# Load packages
library(dplyr)
library(ggplot2)
library(gridExtra)

my_data <- mtcars
glimpse(my_data)
# ?mtcars includes variable definitions

```


## 1. Intro to plotting in base R

Plotting in R can be simple or complicated, depending on your needs. 
To get started, you could use some basic plotting functions, such as 
plot() to create a scatterplot or hist() to create a histogram.
When using these functions, you first specify your data and variables,
and then add additional options related to labels and appearance.

```{r}
# Basic scatterplot
plot(x = my_data$hp, y = my_data$mpg)

# Add a title
plot(x = my_data$hp, y =my_data$mpg, 
     main = "Relationship Between Miles/Gallon and Horse Power")

# Change labels of the axes
plot(x = my_data$hp, y =my_data$mpg, 
     main = "Relationship Between Miles/Gallon and Horse Power",
     xlab = "Horse Power", 
     ylab = "Miles/Gallon")

# Change style of point
plot(x = my_data$hp, y =my_data$mpg, 
     main = "Relationship Between Miles/Gallon and Horse Power",
     xlab = "Horse Power", 
     ylab = "Miles/Gallon",
     pch = 16) 

# Change color of plot
plot(x = my_data$hp, y =my_data$mpg, 
     main = "Relationship Between Miles/Gallon and Horse Power",
     xlab = "Horse Power", 
     ylab = "Miles/Gallon",
     pch = 16,
     col = "purple") 

# Basic histogram (with frequency count)
hist(my_data$mpg, 
     freq = TRUE, 
     main = "Histogram of Miles/Gallon", 
     xlab = "MPG")

# Basic histogram (with relative frequency)
hist(my_data$mpg, 
     freq = FALSE, 
     main = "Histogram of Miles/Gallon", 
     xlab = "MPG")

# Save your plot
# pdf("file_path/plot_name.pdf")
# plot(x = my_data$hp, y = my_data$mpg)
# dev.off()

```

The ggplot2 package is a very powerful tool for creating 
visually pleasing plots. We will use the ggplot() function going
forward. This website is a good resource for learning more:
https://ggplot2.tidyverse.org
Download their "cheatsheet" for ggplot2. 


## 2. Advanced plotting with ggplot: Scatterplots

```{r}

# Basic scatterplot
ggplot(data = my_data, # specify the data
       aes(x = hp, y = mpg)) + # specify the variables (x, y axes)
         geom_point()  # specify how to represent the data points, can add aesthetic options here

# Save the plot as an object and add more aethetic layers
plot.mpg.hp <- ggplot(data = my_data,
                      aes(x = hp, y = mpg)) +
                        geom_point(colour = "purple", shape = 18, size = 3) + # Specify color, shape, and size of points
                        ggtitle("Miles/Gallon vs Horse Power") + # Add title
                        ylab("Miles/Gallon") + # Label y-axis
                        xlab("Horse Power") + # Label x-axis
                        theme(title = element_text(size=12, face = 'bold'), # Increase title font size and make bold
                              plot.title = element_text(hjust = 0.5)) # Center title
  
# Type ?pch in your console to see the corresponding numeric values for different plotting characters. 

```
Now play around with changing the color, size, and other aesethetics. 

## 3. Advanced plotting with ggplot: Histograms

```{r}

# Basic histogram
ggplot(data = my_data, # specify the data
       aes(mpg)) + # specify the variable of interest
        geom_histogram(bins = 8, fill = "blue", alpha = .7) + # Specify number of bins, color, and transparency
        ggtitle("Histogram of Horse Power") + # Add title
        xlab("Horse Power") + # Label x-axis
        ylab("Count") # Label y-axis

# Save the plot as an object
hist.mpg <- ggplot(data = my_data, # specify the data
       aes(mpg)) + # specify the variable of interest
        geom_histogram(bins = 8, fill = "blue", alpha = .7) + # Specify number of bins, color, and transparency
        ggtitle("Histogram of Horse Power") + # Add title
        xlab("Horse Power") + # Label x-axis
        ylab("Count") # Label y-axis

# Export the plot to a PDF
# Note the "~" is a relative file path. It is relative to your R working directory, 
# which can be identified using getwd()
ggsave(filename = "~/AAE 770 - math/R_Materials_AAE636_Zhidong/Bethany/mpg-hp.pdf", # Specify location to save file
       hist.mpg, # Specify plot to save
       device = "pdf")

```

## 4. Plot aesthetics

Let's play around with a few more aethetics. The options are limitless!

```{r}

# Scatterplot: MPG vs weight
# Change color of dots based on manual vs. automatic transmission
scatter1 <- ggplot(data = my_data,
                  aes(x = wt, y = mpg, 
                      color = factor(am, labels = c("Automatic", "Manual")))) + # factor converts this from a continuous variable to a discrete variable. Also add labels for legend to replace 0/1 values
                    geom_point(shape = 18, size = 3) + # Specify shape and size of points
                    xlim(0,6) + # Specify x-axis range
                    ylim(0,40) + # Specify y-axis range
                    labs(title = "MPG vs Weight", # Add title
                         x = "Weight (1,000 pounds)", # Label x-axis
                         y = "Miles per Gallon", # Label y-axis
                         color = "Transmission Type") + # Legend header 
                    theme(title = element_text(size=12, face = 'bold'), # Increase title font size and make bold
                          plot.title = element_text(hjust = 0.5), # Center title
                          legend.position = c(.85, .9)) # Specify legend position

scatter1

# Bar chart: Miles per gallon

# Start by arranging the data by miles per gallon
my_data_2 <- my_data %>%
  mutate(make_model = rownames(my_data))  # Preserve the car make and model info, which is stored as row names


bar1 <- ggplot(my_data_2,
               aes(x = reorder(make_model, mpg), y = mpg, fill = mpg)) + # Sort by mpg, descending
          geom_bar(stat = "identity") + # Specify the height of bars is equal to the data value
          scale_fill_gradient(low = "green", high = "black") + # Specify bar fill colors
          labs(title = "MPG by car make and model",
               x = "Make and Model",
               y = "Miles Per Gallon") +
          theme(legend.position = "none", # Remove legend
                plot.title = element_text(hjust = 0.5)) + # Center title
          coord_flip() # Transpose the plot for horizontal bars

bar1
  

```

## 5. Arranging multiple plots

The gridExtra package allows you to place multiple plots in a
matrix by specifying the number of columns and/or rows.

```{r, message=FALSE}

# install.packages("gridExtra")
gridExtra::grid.arrange(scatter1, bar1, ncol=2)

```

The facet_grid() function allows you to plot data in a matrix, 
based on a discrete variable. Here's an example using the transmission
type.

```{r}

# Scatterplots of MPG vs weight, by transmission type

ggplot(data = my_data,
              aes(x = wt, y = mpg)) + 
                geom_point(shape = 18, size = 3) + # Specify shape and size of points
                xlim(0,6) + # Specify x-axis range
                ylim(0,40) + # Specify y-axis range
                labs(title = "MPG vs Weight", # Add title
                     x = "Weight (1,000 pounds)", # Label x-axis
                     y = "Miles per Gallon") + # Label y-axis
                theme(title = element_text(size=12, face = 'bold'), # Increase title font size and make bold
                      plot.title = element_text(hjust = 0.5)) + # Center title
                facet_grid(cols = vars(factor(am, labels = c("Automatic", "Manual")))) # Specify facet variable and manually label

```